### 2024-08-28 parts/correction.tex
youhouhou

\begin{lemma}
	\label{n-sf-combinatorics:lem}
	Let $f$ be a \kl{star-free $\Nat$-polyregular} function.
	There exists a computable $s \in \Nat_{\geq 1}$
	such that for all \kl{polynomial functions}
	$q \colon \Nat^p \to \Sigma^*$,
	there exists a computable \kl{natural binomial function} $F$
	such that:
	\begin{equation*}
		f \circ q(X_1 + s, \dots, X_p + s)
		=
		F
		\quad
		\text{ over } (\Nat_{\geq 1})^p
		\quad .
	\end{equation*}
\end{lemma}
\begin{proof}[Proof Sketch]
	We redo the proof of \cref{n-poly-combinatorics:lem}, with the extra
	assumption that the monoid $M$ is \kl(monoid){aperiodic}, hence that there
	exists $s \in \Nat$ such that $x^{s+1} = x^s$ for all $x \in M$. As a
	consequence, we can define $\omega = 1$, and
	translate inputs by $s$, instead of multiplying them by $\omega$.
\end{proof}

@@

\AP Let us conclude this section by considering polynomials with
coefficients in $\Rat$, and justify that all the combinatorial work has
already happened in $\Rel$ and $\Nat$. A first observation is that from
\cref{n-poly-combinatorics:lem,n-sf-combinatorics:lem}, we know that the
functions that can be computed by \kl{star-free $\Nat$-polyregular
	functions} are going to be suitable combinations of \kl{natural binomial
	functions}. This motivates the following definition of the ``polynomial
counterpart" of a binomial coefficient, $\intro*\pbinom{X - \ell}{k}$
defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
\intro{binomial monomial}, and introduce \intro{natural binomial
	polynomials} as $\Nat$-linear combinations of products of \kl{binomial
	monomials}, i.e., of the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i
	\prod_{j = 1}^k \pbinom{X_j - p_{i,j}}{k_{i,j}}$. Similarly, we introduce
the class of \intro{integer binomial polynomials}, which are obtained by
$\Rel$-linear combinations of products of \kl{binomial monomials}.

@@

\begin{lemma}
	\label{integer-binomial-polynomial:lem}
	Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial.
	Then, $P$ is an \kl{integer binomial polynomial} if and only if
	$P(\Rel^k) \subseteq \Rel$.
\end{lemma}
\begin{proof}
	Let us first notice that $\pbinom{X}{k}$ is \kl{integer valued}
	when $k \in \Nat$. One way to notice this is that when $X \geq 0$,
	$\pbinom{X}{k} = \binom{X}{k}$ is the number of ways to choose $k$ elements among $X$,
	which is an integer. When
	$X < 0$, then $\pbinom{X}{k} = (-1)^k \pbinom{k - X + 1}{k} = (-1)^k \binom{k - X + 1}{k}$
	is also an integer. Because \kl{integer-valued} polynomials are closed under products
	and $\Rel$-linear combination, we conclude that all \kl{integer binomial polynomials} are
	\kl{integer-valued}.

@@

	Let us now conclude by noticing that if $P$ is \kl{integer-valued},
	then so is $\Delta_i P$ for all $1 \leq i \leq k$. As a consequence,
	so are $\Delta_{\alpha} P$, for all multi-indexes $\alpha$, and we conclude
	that $\Delta_{\alpha}P (0,\dots,0) \in \Rel$ for all $\alpha$.
	Combined with \cref{discrete-taylor:eq}, we conclude that $P$ is an \kl{integer binomial polynomial}.
\end{proof}

@@

\AP We say that a \kl{natural binomial polynomial} is \intro{in base $K$} if
all its \kl{binomial monomials} are of the form $\pbinom{X - K}{d}$ for some $d
	\in \Nat$. To address the issues raised by the previous examples, we introduce
the notion of \intro{strongly natural binomial polynomials} \kl{in base $K$},
as the polynomials $P \in \Rat[X]$ such that for all partial valuation $\nu
	\colon \topartial \set{0, \dots, K}$, $\restr{P}{\nu}$ is a \kl{natural
	binomial polynomial} \kl{in base $K$}. This leads to the following
characterization.

\begin{corollary}
	\label{decide-rat-poly-npoly:cor}
	Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
	coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
	such that $\alpha P \in \Rel[\vec{X}]$. Then, there exists a
	computable $K \in \Nat$, such that the following are equivalent:
	\begin{enumerate}
		\item \label{rat-npoly-1:item}
		      $P$ is a \kl{strongly natural binomial polynomial} \kl{in base $K$},
		\item \label{rat-npoly-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
		\item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
		\item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
		      and for all $\vec{x} \in \Nat^{\vec{X}}$,
		      $P(\vec{x})$ is a multiple of $\alpha$.
	\end{enumerate}
	In particular, the properties are decidable.
\end{corollary}
\begin{proof}
	First, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
	if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
		polynomial}. But in particular, there exists a polyregular function $f$
	that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
	post-compose this function by the one that maps $a^{k \alpha + r}$ to
	$a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
	\kl{represented} by an \kl{$\Nat$-polyregular function}.

	We also know that \cref{rat-npoly-2:item} $\implies$
	\cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
		polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
		\CorrectPoly$ by \cref{corrected-version:thm}.
\end{proof}

>>>>

\begin{lemma}
	\label{n-sf-combinatorics:lem}
	Let $f$ be a \kl{star-free $\Nat$-polyregular} function.
	There exists a computable $s \in \Nat_{\geq 1}$
	such that for all \kl{polynomial functions}
	$q \colon \Nat^p \to \Sigma^*$,
	there exists a computable \kl{natural binomial function} $F$
	such that:
	\begin{equation*}
		f \circ q(X_1 + s, \dots, X_p + s)
		=
		F
		\quad
		\text{ over } (\Nat)^p
		\quad .
	\end{equation*}
\end{lemma}
\begin{proof}[Proof Sketch]
	We redo the proof of \cref{n-poly-combinatorics:lem}, with the extra
	assumption that the monoid $M$ is \kl(monoid){aperiodic}, hence that there
	exists $s \in \Nat$ such that $x^{s+1} = x^s$ for all $x \in M$. As a
	consequence, we can define $\omega = 1$, and
	translate inputs by $s$, instead of multiplying them by $\omega$.
\end{proof}

@@

\AP Let us conclude this section by considering polynomials with coefficients
in $\Rat$, and justify that all the combinatorial work has already happened in
$\Rel$ and $\Nat$. A first observation is that from
\cref{n-poly-combinatorics:lem,n-sf-combinatorics:lem},
we know that the functions that can be computed by \kl{star-free
$\Nat$-polyregular functions} are going to be suitable combinations of
\kl{natural binomial functions}. This motivates the following definition of the
``polynomial counterpart" of a binomial coefficient, given two numbers $\ell,k \in \Nat$,
$\intro*\pbinom{X -
\ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$,\footnote{
    In particular, $\pbinom{X - \ell}{k}$ is defined to be $1$ when $k = 0$, 
    and $X - \ell$ when $k = 1$.
} that we call a
\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
- p_{i,j}}{k_{i,j}}$. Similarly, we introduce the class of \intro{integer
binomial polynomials}, which are obtained by $\Rel$-linear combinations of
products of \kl{binomial monomials}.

@@

\begin{lemma}
	\label{integer-binomial-polynomial:lem}
	Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial.
	Then, $P$ is an \kl{integer binomial polynomial} if and only if
	$P(\Rel^k) \subseteq \Rel$, if and only if $P(\Nat^k) \subseteq \Rel$.
\end{lemma}
\begin{proof}
	Let us first notice that $\pbinom{X}{k}$ is \kl{integer-valued}
	when $k \in \Nat$. One way to notice this is that when $X \geq 0$,
	$\pbinom{X}{k} = \binom{X}{k}$ is the number of ways to choose $k$ elements among $X$,
	which is an integer. When
	$X < 0$, then $\pbinom{X}{k} = (-1)^k \pbinom{k - X + 1}{k} = (-1)^k \binom{k - X + 1}{k}$
	is also an integer. Because \kl{integer-valued} polynomials are closed under products,
	$\Rel$-linear combination, and 
    shifting of domain by an integer,
    we conclude that all \kl{integer binomial polynomials} are
	\kl{integer-valued}.

@@

	Let us now conclude by noticing that if $P$ is \kl{integer-valued},
	then so is $\Delta_i P$ for all $1 \leq i \leq k$. As a consequence,
	so are $\Delta_{\alpha} P$, for all multi-indexes $\alpha$, and we conclude
	that $\Delta_{\alpha}P (0,\dots,0) \in \Rel$ for all $\alpha$.
	Combined with \cref{discrete-taylor:eq}, we conclude that $P$ is an \kl{integer binomial polynomial}.
    Remark that $\Delta_{\alpha} P (0, \dots, 0)$ only ever involves evaluation of $P$ on 
    $\Nat^k$, justifying the weakening of the hypothesis.
\end{proof}

@@

\AP To address the issues raised by the previous examples, we introduce the
notion of \intro{strongly natural binomial polynomials}, as the polynomials $P
\in \Rat[X]$ such that for all partial valuation $\nu \colon \topartial \Nat$,
$\restr{P}{\nu}$ is a \kl{natural binomial polynomial}. This leads to the
following characterization.

\begin{theorem}
	\label{decide-rat-poly-npoly:cor}
	Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
	coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
	such that $\alpha P \in \Rel[\vec{X}]$. Then, 
    the following are equivalent:
	\begin{enumerate}
		\item \label{rat-npoly-1:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
		\item \label{rat-npoly-2:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
		\item \label{rat-npoly-3:item}
		      $P$ is a \kl{strongly natural binomial polynomial},
        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$ 
            and $P$ is \kl{integer-valued}.
	\end{enumerate}
	In particular, the properties are decidable.
\end{theorem}
\begin{proof}

    For the implication \cref{rat-npoly-1:item} $\implies$
    \cref{rat-npoly-4:item}, we obtain $\alpha P \in \CorrectPoly$ via
    \cref{corrected-version:thm} by remarking that the \kl{star-free
    $\Nat$-polyregular functions} have output in $\Nat$ and are closed under
    multiplication by a constant $\alpha \in \Nat$. The fact that $P$ is
    \kl{integer-valued} follows from
    \cref{integer-binomial-polynomials:cor} and the fact that
    $\NSF \subseteq \ZPoly$.

    Then, \cref{rat-npoly-4:item} $\implies$
    \cref{rat-npoly-3:item}, because if $\alpha P \in \CorrectPoly$,
    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ by
    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
    to obtain a number $s \geq 1$ (that only depends on $P$) such
    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
    over $\Nat^k$. Let us write the
    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
    $\card{\alpha} \leq d_2$.
    Now, because of the equality 
    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
    we conclude that we can find a $K \in \Nat$ and coefficients
    $f_{\alpha} \in \Nat$ such that 
    $F(X_1, \dots, X_k) =
    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
    Now, remark that $P$ is an \kl{integer-valued} polynomial of
    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
    \in \Rel$ for all $\card{\alpha} \leq d$
    (\cref{integer-binomial-polynomial:lem}). 
    As a consequence,
    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
    and 
    by identifying coefficients in the previous two expressions, we obtain
    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
    Finally,
    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
    and therefore
    $\lambda P (X_1 + s, \dots, X_k + s) =
    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
    In particular,
    we get
    \begin{equation*}
        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
    \end{equation*}
    And we conclude by term identification
    by term identification
    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
    and $P$
    is a \kl{natural binomial polynomial}.
    Remark that this reasoning remains true whenever we partially apply $P$ to some values
    in $\Nat$, as it amounts to 
    partially applying the function $f$. We have proven that 
    $P$ is a \kl{strongly natural binomial polynomial}.

    Let us now prove by induction on the number of indeterminates that
    \cref{rat-npoly-3:item} $\implies$
    \cref{rat-npoly-2:item}. Note that by construction, there exists a
    number $K \in \Nat$ such that when the input values of $P$ are all greater
    than $K$, $P$ coincides with a \kl{natural binomial function}, which is
    itself \kl{represented} by a \kl{star-free $\Nat$-polyregular function}. If
    some input value $X_i$ is set to a number $x_i \leq K$, then one can
    leverage the fact that $\restr{P}{X_i = x_i}$ remains a \kl{strongly
    natural binomial polynomial} to conclude by induction that $\restr{P}{X_i =
    x_i}$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
    Combining these, we obtain a \kl{star-free $\Nat$-polyregular function}
    \kl{representing} $P$.

    Finally, the implication  \cref{rat-npoly-2:item} $\implies$ \cref{rat-npoly-1:item} is
    immediate as $\NSF \subseteq \NPoly$.
\end{proof}
